---
title: "A Fine Wine...and What Makes It So"
author: "Brian"
date: "April 5, 2017"
output: html_document
---

```{r setup, include=FALSE}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = FALSE)

wine <- read.csv('wineQualityreds.csv')
#install.packages('memisc')
library(reshape2)
library(ggplot2)
library(GGally)
library(gridExtra)
library(psych)
library(dplyr)
library(RCurl)
library(bitops)
library(memisc)
```

## Exploratory Data Analysis 

This analysis will be structured as a stream of consciousness in that the 
majority of it will be graphical looks at the variables behind a red wine 
receiving a high quality rating.  For a more complete and polished summary
of the analysis, please see the bottom three graphs for substantial findings.

American wine making in the last 30 years has exploded and European wine making
has existed for centuries at a very high level.  Behind all the of subjective
decisions a wine maker decides, this dataset shows that there is also reason to
look at the chemical side of when to determine its quality.  This dataset looks
at nearly 1,600 wines and the 12 variables that accompany them.

```{r cars}

wine$X <- NULL
length(wine)
nrow(wine)
summary(wine)


```

## Univariate Plots

One of our most common y variables will be quality.  Here is it's distribution,
looking fairly normal in nature:

```{r pressure}

hist_funct <- function(feature, bins) {
    ggplot(data = wine, aes_string(x = feature)) +
        geom_histogram(binwidth = bins)
}

hist_funct("quality", 1)

#ggplot(wine, aes(x = quality)) +
 # geom_histogram(binwidth = 1)

```

As we'll be hoping to analyze many of the variables, it'll be good to take a 
look at some additional distribtuions:

```{r}

p1 <-hist_funct("volatile.acidity", .05) 

p2 <-hist_funct("citric.acid", 01)

p3 <- hist_funct("residual.sugar", .1)

p4 <- hist_funct("total.sulfur.dioxide", 5)

p5 <- hist_funct("pH", .1)

p6 <- hist_funct("sulphates", .1)

p7 <- hist_funct("alcohol", .1)

grid.arrange(p1, p2, p3, p4, p5, p6, p7, ncol = 4)

```

This approach gives a solid sense of which variables will need to be transformed 
once we start using them as well as a sense of where the majority of wines fall.
Citric acid has a few outliers at 0 that might need to be cleaned up and pH 
levels are all very 
normally distributed. 

I'd like to zoom in on the Citric acid outliers and see if we should remove them:

```{r}
hist_funct("citric.acid", .01) +
  coord_cartesian(xlim = .25)


```

Zooming in you can see there are well over 100 wines with 0 citric acid so I
think it's safe to not consider them false data points.  Further reseach 
has made it clear that wines often have no citric acid in them.  These points
will stay put for now.

I will now examine the residual sugars because of its long tail to the right.
Perhaps it looks different without the outliers

```{r message=FALSE, warning=FALSE}

hist_funct("residual.sugar", .1) +
  xlim(0, quantile(wine$residual.sugar, .90))

summary(wine$residual.sugar)

```

It is far more normal in distribution than it was before.  Looking at its 
summary you can see the max of over 15.  I wonder why some wines have such
extremely high residual sugar levels when so many of them are clustered 
between 1.5 and 3.  

```{r}

hist_funct("sulphates", .01) +
  scale_x_log10() +
  ggtitle('sulphates (log10)') +
  scale_x_continuous(breaks = pretty(wine$sulphates, n = 10))


wine$sulphates.log10 <- log10(wine$sulphates)

```

This was interesting to transform the sulphates and get a more normal 
distribution.  My eye was first drawn to this variable as I've heard people
say that sulphates are the reason people in America get headaches when drinking
red wine but in Europe they don't because they contain fewer sulphates. 
A quick Google search said that was an oversimplification but I'm still 
going to keep my eye on this variable.

```{r message=FALSE, warning=FALSE}

hist_funct("volatile.acidity", .01)+
  scale_x_log10() +
  ggtitle('volatile acidity (log10)') +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 3))

wine$va.log10 <- log10(wine$volatile.acidity)
```
```{r message=FALSE, warning=FALSE}

hist_funct("residual.sugar", .1)+
  coord_trans(x = 'log10')+
  ggtitle('volatile acidity (log10)')

wine$rs.log10 <- log10(wine$residual.sugar)
```

I went ahead and transformed 2 more variables in case it is needed when
building a predictive model.


```{r message=FALSE, warning=FALSE}

summary(wine$alcohol)

```

```{r message=FALSE, warning=FALSE}

ggplot(wine, aes(x = alcohol)) +
  geom_histogram() +
  ggtitle('alcohol')
```


Taking a quick look at alcohol's quantile stats you can see that the min amount
of alcohol in a wine is 8.4 which means 8.4% of the wine's volume.  Alcohol % 
is often discussed when talking about beverages and can often add a bitterness
to a drink it the % is high which is why I'm interested in this variable.  
The max of alcohol is 14.90 which leaves us with a great deal of variance with
this variable and one that might prove to be powerful in prediction.


```{r}

sum(is.na(wine))

```
Before moving on I did a quick check of missing values in this dataset and it's
abnormally complete.  This will make trusting the data a bit easier moving 
forward.

###Univariate Summary:
What is the structure of your dataset?  
There are 1,599 wines in the dataset with 12 features (fixed acidity, volative
acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total 
sulfur dioxide, density, pH, sulphates, alcohol, and quality). All variables
are numeric variables.

Other observations:  

The mean quality rating is 6 on a scale of 1-8 (with no wine receiving below 
a 3)  
Most wines have relatively low residual sugars with a few outliers much higher
than the majority  
The median of pH falls at a 3.3 and it also the most common value  
Alcoholic content in wines is most commonly around 9 and 10, at the 
bottom of the spectrum  

What is/are the main feature(s) of interest in your dataset?  
The main features in the data set are quality, sulphates, and alcohol.  
They are the most known variables and have great variability which I'm
hoping will transfer into predictive power in a model.

What other features in the dataset do you think will help support your
investigation into your feature(s) of interest?  
There is a chance that pH could also play into whether or not a 
high quality score was received seeing as acidity can easily be tasted 
in comparison to other variables such as density.

Did you create any new variables from existing variables in the dataset?  
I created two subsetted dataframes, one for dry wines and one for sweet wines. 
My hope was that they would lead to quantitatively different quality ratings
but their means are very close to each other.

Of the features you investigated, were there any unusual distributions?
Did you perform any operations on the data to tidy, adjust, or change the 
form of the data? If so, why did you do this?
There were a few unusual distributions that I tried to normalize so that 
they could be more easily used in a regression model later on.  I transformed
sulphates with a log10 so that it went from being skewed to normal.  I
attempted to normalize alcohol but neither of the log10, sqrt, or reverse
transformations truly made a difference.


##Bivariate Analysis

```{r message=FALSE, warning=FALSE}
wine.dry <- subset(wine, residual.sugar < 2.2)

wine.sweet <- subset(wine, residual.sugar > 2.2)

ggplot(wine.dry, aes(x = residual.sugar, y = quality)) +
  geom_jitter()

ggplot(wine.sweet, aes(x = residual.sugar, y = quality)) +
  geom_jitter()

summary(wine$quality)
summary(wine.dry$quality)
summary(wine.sweet$quality)

```

I wanted to take a look a the different on quality between'sweet' and 'dry'
wines.  This seems to be determined by the amount of residual sugars left in 
the wine after the alcohol has fermented.  To do so I created 2 new dataframes
for each category of wine.  The median amount was 2.2 and so I classified wines
with more than 2.2 as sweet and those with less as dry.  The  
mean difference between the two groups was only .07 and without a significance
test, I'm not sure it really tells us anything.  A dead end, but that's not
uncommon in EDA.  

```{r   message=FALSE, warning=FALSE}
#mat <- cor(wine)
#round(mat, 2)

ggcorr(wine)

```

The above correlation plot allows us to very quickly spot variables that
either have a positive or negative correation to each other.  Several of these
are only correlated because they are transformations of each other.  
Another type of visualization to see how variables interact is a mixed plot
matrix.  Let's take a look at the inter-correlations below:

```{r echo = FALSE}
#ggpairs(wine, mapping = aes('psych'))

wine$quality <- as.factor(wine$quality)

p_ <- function(pm) {
  if (interactive()) {
    print(pm)
  }
  invisible()
}

#data(tips, package = "reshape")
pm <- ggpairs(wine, c(3,10,12,14))
p_(pm)
#pm <- ggpairs(wine, upper = "blank")
#p_(pm)

pm
```

This plot shows that our inter-correlation of the independent variables is quite
weak, which is a good thing for future modeling.  It also highlights the fact 
that our distribution isn't too skewed except for alcohol.


```{r   message=FALSE, warning=FALSE}

ggplot(wine, aes(x = fixed.acidity, y = pH)) +
  geom_point()

```

```{r   message=FALSE, warning=FALSE}
cor.test(wine$fixed.acidity, wine$pH)
```


Here we see that two of our variables do have a strong interaction.  Fixed 
acidity and pH which makes a lot of sense seeing as pH is our scale of how acid
or base a substance is.  It is always good to see assumptions about variable 
shown by statistical tests.

```{r   message=FALSE, warning=FALSE}

ggplot(wine, aes(x=density, y = fixed.acidity)) + 
  geom_point()
```

The density of wine seems fairly straightforward in reseach, the weight of
the liquid by volume.  It seems to be affected by a few ingredients such as 
alcohol, sugars, etc.  Here we see that it's either affecting the acidity or
vice versa!

```{r   message=FALSE, warning=FALSE}
ggplot(wine, aes(x=density, y = fixed.acidity)) + 
  geom_point(alpha = 1/3)
```

After adjusting the plot to take a closer look at concentrations of the data
points, you can see that a majority of the density data is between 0.995 and
1.000.  And yet even in that tightly clustered range you can see the positive
correlation with fixed.acidity.

```{r echo = FALSE}


ggplot(wine, aes(x=density, y = residual.sugar)) + 
  geom_point()

```

```{r echo = FALSE}
ggplot(wine, aes(x=density, y = residual.sugar)) + 
  geom_point(alpha = 1/5)
```

Here again you can see the clustering of the density data but this time with a
much weaker correlation to residual.sugar.  These variables with 
intercorrelation are helpful to view because they start to paint a picture 
of which variables might start to affect a regression model.  
It's important to remember that highly correlated variables should not both be
used in a linear regression model because that can cause undesired noise and
interference.

```{r   message=FALSE, warning=FALSE}

wine$quality <- as.factor(wine$quality)

ggplot(wine, aes(x=quality, y = alcohol)) + 
  geom_jitter() + 
  geom_boxplot(alpha = .5) +
  stat_summary(color = 'red')

```

THe above graph is the first between an independent variable and the dependent
variable of quality.  The correlation between the two was .48 and so I expect
to be using it to predict wine quality ratings.  It is a little more difficult
in the above graph to visualize relatedness because of the categorical nature
of the dependent variable, but you are able to see that at max end of the 
quality spectrum there is a higher positive correlation.  The boxplots are able
to higlight this relationship along with the mean summary points.

```{r   message=FALSE, warning=FALSE}
wine$quality <- as.numeric(wine$quality)

ggplot(wine, aes(x=alcohol, y = quality)) + 
  geom_point() +
  #geom_jitter(position = position_jitter()) +
  geom_jitter(width = 2, height = 1.5) +
  geom_smooth(method = lm, color = 'green')
```

To help get a better picture of quality and alcohol's relationship despite the
difficulty with dealing with a categorical variable, I jittered the plot and 
added a linear best fit line.  You can clearly see the positive relationship
between the two.

```{r echo=FALSE}
quality_groups <- group_by(wine, quality)

wine.quality_groups <- summarize(quality_groups,
                          alcohol_mean = mean(alcohol),
                          pH_mean = mean(pH),
                          sulphates_mean = mean(sulphates),
                          n = n())

wine.quality_groups <- arrange(wine.quality_groups, quality)

head(wine.quality_groups)

#summary(wine.quality_groups)
```

Grouping by quality gave me a clear look at how a few different variables (at 
least their averages) interacted with the dependent variable.  The graphs
above also accomplished this, but sometimes it's a bit clearer to see mean 
comparisons in numerical format side by side.


###Bivariate Summary
####Talk about some of the relationships you observed in this part of the
####investigation. How did the feature(s) of interest vary with other features in
####the dataset?
An interestingly strong correlation was between quality and alcohol at 0.48

The second strongest correlatoin with quality was a negative correlation
at -0.39

Density seemed to correlate strongly with a few different variables with one
of them being acidity. 

The intercorrelation between the independent variables were all considerably
weak as demonstrated in the plot matrix above.  

Alcohol had a skewed distribution so I transformed it because of its potential
power in future predictions due to its strong correlation with quality

####Did you observe any interesting relationships between the other features
I'm still fascinated a bit by density and its relationship with many different
variables.  It has high correlations with volatile acidity and alcohol and 
numerous moderate strength correlations to other variables.  It would seem as 
though the density of a wine truly depends on several factors.

####What was the strongest relationship you found?
The strongest relationship found was between pH and fixed acidity with a 
Pearson's r value of 0.68


##Multivariate

```{r   message=FALSE, warning=FALSE}

wine$quality <- as.factor(wine$quality)

p_ <- function(pm) {
  if (interactive()) {
    print(pm)
  }
  invisible()
}

#data(tips, package = "reshape")
pm <- ggpairs(wine, c(3,5,9,10,11,12), 
              mapping=ggplot2::aes(colour = quality),
              legend = c(3,3))
              
#p_(pm)
#pm <- ggpairs(wine, upper = "blank")
#p_(pm)

pm
```

The plot matrix above adds an additional layer of detail when comparing 
the characteristics of wine with its quality rating.  For instance, you can
see that the filled in line graph for alcohol shows that many lower alcohol
%s received the quality rating of 5.  This is why the correlation is so storng
between the two.

```{r   message=FALSE, warning=FALSE}
ggplot(wine, aes(x = alcohol, y = volatile.acidity)) +
  geom_point() + 
  geom_jitter(position = position_jitter(), aes(color = quality)) +
  scale_color_brewer(palette = "Greens") +
  theme_dark()
  #geom_jitter(aes(color = quality))
```

Seeing how strongly both volatile acidity and alcohol interacted with quality
in the correlation matrix in the bivariate analysis, I wanted to see if both 
interplayed at all with quality.  This scatter plot overlayed with quality
as color shows us that with the heavy green color on the left and heavy light
and dark blue on the right, that quality is dependent on these two variables.
Because there is little variation though in the vertical sense, it also 
seems to point that there quality is more heavily affected by alcohol and that
volatile acidity simply seems to be lower with higher alcohol percentages.


```{r   message=FALSE, warning=FALSE}
wine$quality <- as.numeric(wine$quality)

attach(wine)
regr1 <-lm(formula=quality~alcohol+chlorides+citric.acid+density+fixed.acidity+free.sulfur.dioxide+pH+rs.log10+total.sulfur.dioxide+sulphates.log10+va.log10)

regr1
```

To start our linear modeling attempt, I simply add all of our variables into
the model because none of them correlate strongly enough to add considerable 
noise.  

```{r   message=FALSE, warning=FALSE}
summary(regr1)
```

Here we take a look at the model output.  The R squared (strength) of the model
is 0.368 which is not very powerful.  That is to be considered give the fact 
that we simply through all of our variables (plus one transformation) into it.
The power of a model comes when unecessary variables are taken out.  By looking
at the p values I narrowed down the model coefficients for round 2 below.

```{r   message=FALSE, warning=FALSE}
regr2 <-lm(formula=quality~alcohol+chlorides+volatile.acidity+sulphates.log10
           +pH+total.sulfur.dioxide+rs.log10+free.sulfur.dioxide+citric.acid)

summary(regr2)
```

This time the model is able to predict over 37% of red wine qualities.  This is 
not a considerably powerful model nor a marked improvement over the first model
with its predictive power of 36%.  Further, more specific transformations of
variables along with a great knowledge of wine industry could considerably
improve this attempt.

```{r}
par(mfrow=c(2,2))
plot(regr2)
```

This plot of our residuals from the linear regression model allows us to do a
quick spot check.  Although the residuals graph in the upper left looks different
than most because of the categorical nature of the dependent variables, one still
can see that it is roughly a normal distribution.  


###Multivariate Summary

####Talk about some of the relationships you observed in this part of the 
####investigation. Were there features that strengthened each other in terms of
####looking at your feature(s) of interest?
The relationship between alcohol content and quality continued to be 
strengthened by additional plots.  Even when weaving in volatile acidity,
the correlation seemed to point more towards alcohol doing the work in adjusting
quality. 

####Were there any interesting or surprising interactions between features?
I was suprised to see that multiple variables dealing with sulphur ended up
being significant in the linear regression model that I ran.  I expected them 
to be strongly correlated to each other but because that was not the case
they did not seem to cause multiple collinearity in the model.

####Did you create any models with your dataset? Discuss the
####strengths and limitations of your model.
I ran 2 linear regression models with the dataset both having the dependent
variable be quality.  The first simply included
all the variables and had an R squared value of .36  Being able to only predict
36% of the data is not very strong.  After whittling down to just the 
significant variables, the model rose to over 37% predicitve power.  Still not
impressive which means a lot more work can be done with the dataset.

##Final Plots and Summary
###Plot 1

```{r   message=FALSE, warning=FALSE}
ggplot(wine, aes(sulphates)) + 
  geom_histogram(binwidth = .01) +
  scale_x_log10() +
  ggtitle('Log 10 of Sulphate')+
  ylab('Number of Wines')+
  xlab('Sulphate Levels (potassium sulphate - g / dm3)') +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))
```
```{r warning=FALSE, message = FALSE}

wine$quality <- as.factor(wine$quality)

ggplot(wine, aes(x=quality, y = sulphates.log10)) + 
  geom_jitter() + 
  geom_boxplot(alpha = .5) +
  ylab('Sulphate Levels Transformed log10')+
  xlab('Quality as factor')+
  stat_summary(color = 'red')
```


###Description 1
The distribution of the sulphate variable is very skewed before transformation.
This graph is a log10 transformation of it because of its significance in 
regards to determing the quality of a red wine.  This significan is shown in the
boxplot portion where you can see the positive correlation between sulphate 
levels and wine quality ratings.

##Plot 2

```{r   message=FALSE, warning=FALSE}

wine$quality <- as.factor(wine$quality)

ggplot(wine, aes(x = alcohol, color = quality, fill = quality)) +
  geom_density(alpha = 0.5) +
  ggtitle('Alcohol % by Quality')+
  xlab('Alcohol (% by volume)') +
  ylab('Percent of total') +
  scale_fill_brewer(palette = "Greens") +
  scale_color_brewer(palette = "Greens") +
  theme_dark()
  

```

###Description Plot 2 
The second plot is an extension of the plot matrix in the Multivariate Analysis
section.  It is an easy way to identify the relationship between alcohol
and quality as you can see that the majority of wines with a quality of 3
are below 10% alcohol while many of the 5s and 6s are above 11.

##Plot 3

```{r   message=FALSE, warning=FALSE}
ggplot(wine, aes(x = alcohol, y = volatile.acidity)) +
  geom_point() + 
  geom_jitter(position = position_jitter(), aes(color = quality))+
  xlim(9,14) +
  ylim(.1, 1) +
  ggtitle('Volatile Acidity Comparison with Alcohol')+
  xlab('Alocohol (% by volume)') +
  ylab('Volatile Acidity') +
  scale_color_brewer(palette = 'Greens')
```

### Description Plot 3
This plot shows that both volatile acidity and alcohol play a part in the
quality of a wine, but that alcohol is a stronger relationship.  In addition,
wines that have more alcohol tend to have a lower volatile acidity.  The 
correlation between volatile acidity and alcohol is -0.2 but there obviously
are still findings within the graph that are worthwhile.  The fact that alcohol 
is positively correlated with quality and volatile acidity is negatively
correlated to quality is concisely shown in one graph.

##Reflection
The red wine quality dataset that I explored has nearly 1,600 datapoints to 
examine spread out across 12 different variables.  The majority of the variables
were numeric with one being factorl, the variable of quality. I first made 
several histograms to get a feel for the distribution of the data and then 
performed a few transformations on it to aide in eventually fitting it to a 
linear regression model.  The data was very complete and did not take additional
cleaning.

I placed wines into 2 groups, one for sweet wines and one for dry wines.  This
grouping did not seem to show any strong differences between the two.  I also
group the wines based on their quality datapoint and looked at the differences
in means of variables between them.  This was helpful confirmation when deciding
on most powerful explanatory variables.

I was unable to do as much intervariable analysis because so many of the 
variables were numeric instead of factoral.  The fact that quality was
factoral did help in the end as I examined a few plots with the extra layer of
quality on top.

Several variables stood out very quickly in terms of being highly correlated 
with quality.  A correlation matrix was very helpful in determining those
datapoints.  Alcohol, volatile acidity, and sulphates were all correlated with
quality and ended up being variables in the regression model.  Suprisingly so,
there were a few variables that also contributed to the model that I did not
expect such as chlorides.  All told, though, the model was not very powerful
and was able to only explain about 37% of the datapoints.

The dataset does have its downsides with one of them being only one quality
rating.  If the point of collecting so many different wines and their properties
is to see which elements affect the rating they receive, it would be very 
helpful to have 5 or more ratings and then to take an average of the spread. 
Without this, one has to assume that the only quality rating a wine received
is the absolute truth for that wine.  Lastly, the region breakdown of the wines
would be helpful to examine. One place I read online said that this dataset
all came from Portugal.  If that is the case, one could examine which wines
have properties that are consistent with others in their area and if certain
areas receive higher quality rankings than others.

